---
title: "NYC Green Taxis Analysis_C230522"
output: html_notebook
---


Calling the required libraries  
```{r}
library(data.table)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(randomForest)
library(mltools)
```

Import the data set
```{r}
###### Read the data set #####
data <- fread("green_tripdata_2015-09.csv", stringsAsFactors = F)
head(data)
```
Dimension of data set
```{r}
dim(data)
```
Summary statistics of the data set
```{r}
summary(data)
```
To check if there any null values:
Ehail the whole column has null value, so dropping that column
Trip_Type has 4 missing values, so replacing the nas with the most frequent value
```{r}
summary(is.na(data))
```
```{r}
data$Ehail_fee<-NULL
a<-data.frame(table(data$Trip_type))
data$Trip_type[which(is.na(data$Trip_type)==T)] <- a$Var1[which.max(a$Freq)]
summary(is.na(data$Trip_type))
```

Frequency of Distance Travelled seems to be Rightly skewed
```{r}
his<-ggplot(data, aes(x=Trip_distance)) + 
  geom_histogram(color="black", fill="red", bins=100 ) + labs("Histogram of Trip duration") +
  ylab("Frequency") + xlab("Total Distance")
his
```
Checking the frequency of Trip distance for distance < 20  seems better than the previous histogram indicating that there are outlier values
```{r}
his<-ggplot(data[data$Trip_distance<200,], aes(x=Trip_distance)) + 
  geom_histogram(color="black", fill="red",bins=100 ) + labs("Histogram of Trip duration") +
  ylab("Frequency") + xlab("Total Distance")
his
```

Histogram after taking log of Trip Distance since the data is skewed
```{r}
his<-ggplot(data, aes(x=log(Trip_distance))) + 
  geom_histogram(color="black", fill="red" ) + labs("Histogram of Trip duration") +
  ylab("Frequency") + xlab("Total Distance")
his
```
```{r}
box<- ggplot(data, aes(x="",y=Trip_distance)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=5)
box
```
Summary of Trip Distance suggests that even though the mean is just 2.968, there is some outlier which makes the max value very high.
```{r}
summary(data$Trip_distance)
```
Creating useful variables by extrcting the hour of day, week,month and weekday
```{r}
data[,pickup_hour:=hour(lpep_pickup_datetime)]
data[,pickup_week := week(lpep_pickup_datetime)]
data[,pickup_month := month(lpep_pickup_datetime)]

data[,pickup_weekdays := weekdays(as.POSIXct(data$lpep_pickup_datetime,abbreviate=F))]

data[,pickup_datetime:=as.Date(lpep_pickup_datetime)]
data[,dropoff_datetime:=as.Date(Lpep_dropoff_datetime)]
data[,time_diff:= (as.POSIXct(Lpep_dropoff_datetime)-as.POSIXct(lpep_pickup_datetime))/60]

```

Mean and Median of Trip Distance across hours of the day
Both the mean and median of the trip duration is at its peak in the morning 5 to 6 and is lesser during the day

```{r}
stat_duration <-data[,list(mean_trip=mean(Trip_distance),
                           median_trip=median(Trip_distance)), by=list(pickup_hour)]

mean_trip_plot <- ggplot(stat_duration,aes(x = as.factor(pickup_hour), y = mean_trip)) +
  geom_bar(stat = 'identity', fill = 'red') +
  labs(x = 'Hours', y = 'Mean Trip Duration', title = 'Mean Trip duration by hour of the day') 

median_trip_plot <-ggplot(stat_duration,aes(x = as.factor(pickup_hour), y = median_trip)) +
  geom_bar(stat = 'identity', fill = 'red') +
  labs(x = 'Hours', y = 'Median Trip Duration', title = 'Median Trip duration by hour of the day')


grid.arrange(mean_trip_plot, median_trip_plot, nrow=2)
```

Subsetting the data for airport using RateCode. RateCode =2 is JFK, Ratecode =3 is Newark

```{r}
airport <- subset(data,data$RateCodeID == 2 | data$RateCodeID ==3)

air <- airport[,list(number= length(VendorID), Average_Fare = mean(Fare_amount), 
                     average_passengers = mean(Passenger_count)),
               by =list(RateCodeID)]

```

The trip distance to or fro the airports are mostly more than other trips throughout the day except for 11:00 PM to 1:00 AM where the trip duration is more for the non airport regions. This is inspite of the fact the the number of non-airport rides is much higher than the airpoirt rides

```{r}
temp<-data


temp[,air_flag:=ifelse(RateCodeID==2 | RateCodeID ==3,'Airport','Non Airport')]

air_noair <-temp[,list(avg=mean(time_diff)),by=list(air_flag,pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = avg, fill=air_flag)) +
  geom_bar(stat = 'identity', position=position_dodge(), color = "black") + 
  labs(x = 'Hours', y = 'Mean Trip Duration ', title = 'Mean Trip duration by hour of the day for airports vs non airport rides')


air_noair1 <-temp[,list(count=length(time_diff)),by=list(air_flag,pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = count, fill=air_flag)) +
  geom_bar(stat = 'identity', position=position_dodge(), color = "black") + 
  labs(x = 'Hours', y = 'Number of trips ', title = 'Number of trips by our of the day  for airports vs non airport rides')
grid.arrange(air_noair1, air_noair, nrow=2)

```

Even though airport rides make up to only 0.37% of total rides, the fare from airport rides is about 1.41%. The average cost per ride for airports is $57.20 while for Non-Airport rides is $14.87

```{r}
comp<-temp[,list(No_of_ride=length(Total_amount),Total_Fare=sum(Total_amount),Total_Tip=sum(Tip_amount)),by=list(air_flag)]
comp<- mutate(comp, No_of_ride_percent = No_of_ride / sum(No_of_ride), Total_Fare_percent= Total_Fare/sum(Total_Fare), avg_cost_per_ride = Total_Fare/No_of_ride)
comp
```

Mean trip Duration for both the airports together

```{r}
both_airport <-airport[,list(avg=mean(time_diff)),by=list(pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = avg), group =1) +
  geom_bar(stat = 'identity', fill = 'green') + geom_line(colour = 'red') + 
  geom_point(colour='red')+
  labs(x = 'Hours', y = 'Mean Trip Duration ', title = 'Mean Trip duration by hour of the day for both the airports')
both_airport

```

The mean trip duration for Newark is highest from 2:00 AM to 4L00 AM while for JFK , the trip duration is high almost throughout the day

```{r}
jfk<- airport[,list(avg=mean(time_diff[RateCodeID==2])),by=list(pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = avg), group =1) +
  geom_bar(stat = 'identity', fill = 'green') + geom_line(colour = 'red') + 
  geom_point(colour='red')+
  labs(x = 'Hours', y = 'Mean Trip Duration', title = 'Mean Trip duration by hour of the day for JFK')

new <-airport[,list(avg=mean(time_diff[RateCodeID==3])),by=list(pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = avg), group=1) +
  geom_bar(stat = 'identity', fill = 'green') + geom_line(colour = 'red') + 
  geom_point(colour='red')+
  labs(x = 'Hours', y = 'Mean Trip Duration', title = 'Mean Trip duration by hour of the day for Newark')
grid.arrange(jfk, new, nrow=2)


```

The number of passengers for JFK keeps on increasing till 3:00 PM and starts declining showing that more flights from JFK depart during the day.
This trend is simillar for Newark

```{r}
jfk<- airport[,list(avg=sum(Passenger_count[RateCodeID==2])),by=list(pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = avg), group =1) +
  geom_bar(stat = 'identity', fill = 'green') + geom_line(colour = 'red') + 
  geom_point(colour='red')+
  labs(x = 'Hours', y = 'Total Passenger Count', title = 'Total Passenger Count by hour of the day for JFK')

new <-airport[,list(avg=sum(Passenger_count[RateCodeID==3])),by=list(pickup_hour)]%>%
  ggplot(aes(x = (pickup_hour), y = avg), group=1) +
  geom_bar(stat = 'identity', fill = 'green') + geom_line(colour = 'red') + 
  geom_point(colour='red')+
  labs(x = 'Hours', y = 'Total Passenger Count', title = 'Total Passenger Count by hour of the day for Newark')
grid.arrange(jfk, new, nrow=2)

```

More number of passenegers travel to or fro JFK during Friday. While for Newark, most of the rides are on Sunday and Tuesday.
This might suggest that people use JFK for more of holiday travels and Newark for short business trips


```{r}
jfk<- airport[,list(avg=sum(Passenger_count[RateCodeID==2])),by=list(pickup_weekdays)]%>%
  ggplot(aes(x = factor(pickup_weekdays,levels = c("Friday", "Saturday", "Sunday", "Monday","Tuesday","Wednesday","Thursday")), y = avg)) +
  geom_bar(stat = 'identity', fill = 'green')+
  labs(x = 'weeks', y = 'Total Passengers', title = 'Total Passenger Count by week of the day for JFK')



new<- airport[,list(avg=sum(Passenger_count[RateCodeID==3])),by=list(pickup_weekdays)]%>%
  ggplot(aes(x = factor(pickup_weekdays,levels = c("Friday", "Saturday", "Sunday", "Monday","Tuesday","Wednesday","Thursday")), y = avg)) +
  geom_bar(stat = 'identity', fill = 'green')+
  labs(x = 'weeks', y = 'Total Passengers', title = 'Total Passenger Count by week of the day for Newark')


grid.arrange(jfk, new, nrow=2)

```

Even though the number of rides to JFK is more, but the average fare per ride is more for Newark

```{r}
airport[,air_flag:= ifelse(RateCodeID ==2,'JFK','Newark')]
comp<-airport[,list(No_of_ride=length(Total_amount),Total_Fare=sum(Total_amount),Total_Tip=sum(Tip_amount)),by=list(air_flag)]
comp<- mutate(comp, No_of_ride_percent = No_of_ride / sum(No_of_ride), Total_Fare_percent= Total_Fare/sum(Total_Fare), avg_cost_per_ride = Total_Fare/No_of_ride)
comp
```
```{r}
data[,tip_per:=Tip_amount/Total_amount]
### to cinvert into hour
data[,time_diff:=time_diff/ (60)]

```

```{r}
data_temp1<- subset(data,is.na(data$tip_per) ==F)
  
```

Dropping the unwanted columns

```{r}
data_temp1$lpep_pickup_datetime<- NULL
data_temp1$Lpep_dropoff_datetime<-NULL
data_temp1$VendorID<-NULL
data_temp1$Store_and_fwd_flag<-NULL
data_temp1$Tip_amount<-NULL
data_temp1$pickup_week<-NULL
data_temp1$pickup_month<-NULL
data_temp1$pickup_datetime<-NULL
data_temp1$dropoff_datetime<-NULL
data_temp1$Pickup_longitude<- NULL
data_temp1$Pickup_latitude<-NULL
data_temp1$Dropoff_latitude<-NULL
data_temp1$Dropoff_longitude<-NULL

```

Converting the categorical columns into factors

```{r}
data_temp1$RateCodeID<- factor(data_temp1$RateCodeID)
data_temp1$pickup_hour<- factor(data_temp1$pickup_hour)
data_temp1$Payment_type<- factor(data_temp1$Payment_type)
data_temp1$Trip_type<- factor(data_temp1$Trip_type)
data_temp1$pickup_weekdays<- factor(data_temp1$pickup_weekdays)

```

Creating one hot vector for categorical columns and dropping the base column

```{r}
data_temp1<-one_hot(data_temp1,dropCols = TRUE)
data_temp1$RateCodeID_1<-NULL
data_temp1$pickup_hour_0<-NULL
data_temp1$Payment_type_1<-NULL
data_temp1$Trip_type_1<-NULL
data_temp1$pickup_weekdays_Friday<-NULL
```

Dividing the data into train and test 70:30
Ran Regression model
The adjusted R square being 78.55%.
Most of the variables are significant. 
But Trip_type is not significant at all.
So removing this variable to re-run the model.

```{r}
smp_size <- floor(0.7 * nrow(data_temp1))
set.seed(2018)
train_ind <- sample(seq_len(nrow(data_temp1)), size = smp_size)

train <- data_temp1[train_ind, ]
test <- data_temp1[-train_ind, ]


fit = lm(tip_per ~., train )
summary(fit)

```

Removing Trip_type2 and rerunning the model.
The adjusted R square being 78.55%.
Hence, 78.55% variation in tip_per is due to variation in RateCodeID,Passenger_Count, Trip_Distance,Fare_Amount, Extra,MTA_Tax, Tolls_Amount,improvement_surcharge , Payment_type,Total_Amount,Pickup_hour,Pickup_weekday, time_diff
Most of the variables are signifiacnt as well


```{r}
data_temp1$Trip_type_2<-NULL
smp_size <- floor(0.7 * nrow(data_temp1))
set.seed(2018)
train_ind <- sample(seq_len(nrow(data_temp1)), size = smp_size)

train <- data_temp1[train_ind, ]
test <- data_temp1[-train_ind, ]


fit = lm(tip_per ~., train )
summary(fit)

```

Predicting for test, the MSE =0.1677%
This is a good model

```{r}
test1<- test
test1$tip_per<-NULL
pred <- predict.lm(fit,test1)

(mean((test$tip_per - pred)^2)*100)
```

Taking a sample of 10000 since my computer is unable to process random forest regressors with the full data

```{r}
temp <- sample_n(data, 10000)
temp$Pickup_longitude<-NULL
temp$Pickup_latitude<-NULL
temp$Dropoff_longitude<-NULL
temp$Dropoff_latitude<-NULL
temp$lpep_pickup_datetime<-NULL
temp$Lpep_dropoff_datetime <-NULL
temp$Tip_amount <-NULL
data_temp1<- subset(temp,is.na(temp$tip_per) ==F)

```

Converting the characters to factors

```{r}
data_temp1<- data_temp1%>% mutate_if(is.character, as.factor)
```

Getting train and test data

```{r}
smp_size <- floor(0.7 * nrow(data_temp1))

set.seed(2018)
train_ind <- sample(seq_len(nrow(data_temp1)), size = smp_size)

train <- data_temp1[train_ind, ]
test <- data_temp1[-train_ind, ]


```

Running the random forest 
The MSE is 0.110%
The % of variance in tip_per explained by the variables is 87.03% which is more than the regression shown above

```{r}
model <- randomForest(tip_per~.,data=train, importance=T)
model
```

This shows which variables are important: Trip_Distance, Time_diff, Fare_amount, Total_amount, Payment_type

```{r}
# Variable Importance Plot
varImpPlot(model,
           sort = T,
           main="Variable Importance",
           n.var=10)
```

Predicting the test data, the MSE is 0.1237%

```{r}
test1<- test
test1$tip_per<-NULL

pred <- predict(model,test1)


(mean((test$tip_per - pred)^2))*100
```

Conclusion for prediction of tip_per:
The random forests can be run for the full data. Based in my analysis here
Regression- Variation Explained: 78.55%, MSE: 0.1677%
Random Forest- Variation Explained: 87.03%, MSE: 0.1237%

Hence, Random Forest is a better model



Variable for Average Speed column

```{r}
data[,time_diff:=as.numeric(time_diff)]
data[,avg_speed:=(Trip_distance/time_diff)]
```

Weekly avearge speed acoss the weeks of September shows that the average speed is highest on week number 39

```{r}
ano_data<- data[,c('avg_speed','pickup_week')]
ano_data<- subset(ano_data,abs(ano_data$avg_speed)!=Inf)
ano_av<- ano_data[,list(avg=mean(avg_speed)), by=list(pickup_week)]
ggplot(ano_av, aes(x = pickup_week, y = avg)) +
  geom_bar( stat = "identity", aes(fill = factor(pickup_week))) +
  labs(title =" Weekly Average Speed in September 2015", 
       x = "Week Number of the year 2015",
       y = "Weekly Average Speed") +
  scale_fill_discrete(name = "Week number")

```

Anova test to check if the average speed is same through the weeks of september.
Null hypothesis: states that all the average speeds across the weeks are equal
Alternative hypothesis : Atleast one or all the average speeds across the weeks are unqueal
Since the p value 0.00111 < 0.05 , we reject the Null Hypothesis. 
Hence, the average speeds across the weeks are uneuqal

```{r}

ano <- aov(avg_speed~as.factor(pickup_week), data = ano_data)
summary(ano)
```

To check the pairwise difference in mean. looking at the p value suggests that the pairs 39-38 and 39-37 show statistical difference in mean since their p value < 0.05

```{r}
tuk <-TukeyHSD(ano)
tuk
plot(tuk)
```


```{r}
box <- ggplot(ano_data,aes(x=factor(pickup_week), y=avg_speed)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=5) +   labs(x = "Week Number of the year 2015",
               y = "Average Speed")

box  
```

Anova test to check if the average speed is same through the hours of the day.
Null hypothesis: states that all the average speeds through the hours of the day are equal
Alternative hypothesis : Atleast one or all the average speeds across the hours of the are unqueal
Since the p value <2e-16 < 0.05 , we reject the Null Hypothesis. 
Hence, the average speeds across the hours are uneuqal

```{r}
ano_data<- data[,c('avg_speed','pickup_hour')]
ano_data<- subset(ano_data,ano_data$avg_speed!=Inf)
ano <- aov(avg_speed~as.factor(pickup_hour), data = ano_data)
summary(ano)
```

To check the pairwise difference in mean. looking at the p value suggests there are many pairs such as 5-1 , 16-2 etc. show statistical difference in mean since their p value < 0.05

```{r}
tuk <-TukeyHSD(ano)
tuk
plot(tuk)
```

The avearge hourly speed is maximum at 05:00 AM and declines as the day progresses and again increases in the evening after 6:00 PM

```{r}
ano_av<- ano_data[,list(avg=mean(avg_speed)), by=list(pickup_hour)]

ggplot(ano_av, aes(x = pickup_hour, y = avg)) +
  geom_bar( stat = "identity", aes(fill = factor(pickup_hour))) +
  labs(title ="Hourly Average Speed in September 2015", 
       x = "Hour of the day",
       y = "Hourly Average Speed") +
  scale_fill_discrete(name = "Hour")
```

```{r}

box <- ggplot(ano_data,aes(x=factor(pickup_hour), y=avg_speed)) + 
  geom_boxplot(outlier.colour="red", outlier.shape=5) +   labs(x = "Hour of the day",
                                                               y = "Average Speed")

box               

```

